Visualizing demographic evolution using geographically inconsistent census data

### 2018-08-16 paper/discussion.tex
beta?

The objective of this work was to leverage a graph based data representation
with visual tools to allow for the exploration of geographically inconsistent
census data. While we successfully replicated and corroborated results from the
literature, this method still has significant limitations.

Removing the geographical normalization/interpolation step greatly reduces the
amount of work necessary, but the method still requires consistent variables
across the years. Matching the fields of the public census can be trivial for
some aspects (Age), but challenging for others (Income). The divulged income
ranges vary over time, the actual value changes due to inflation and other
factors, and so on. Moreover, some fields were not considered in earlier
censuses, such as Race in Canada, or Hispanic population in the USA, hampering
its use when they are available. Since this is only a prototype, we matched few
aspects, but a deeper demographic analysis would greatly benefit from all
available information.

>>>>

Our objective was to leverage a graph based data representation and
visualization methods for the exploration of geographically inconsistent
region-based data. While we successfully replicated and corroborated results
from the literature, this method still has significant limitations.

Removing the need for geographical harmonization greatly reduces the amount of
work necessary to explore demographic data, but the method still requires
consistent variables across the years. Matching the variables can be trivial for
some aspects (Age), but challenging for others (Income). The divulged income
ranges vary over time and the actual values change due to inflation. Some
variables were not considered in earlier censuses, such as Race in Canada, or
Hispanic population in the USA, hampering its use when they are available. Since
this is only a prototype, we matched few aspects, but a proper demographic
analysis would benefit from all available information.

While using one small map for each year leads to an easier visualization that
does not require interaction, it does not scale if more than five or six years
are considered. In this case, it might be interesting to replace the larger map
considering each year individually, along with a temporal control for
navigation. Indeed, including more years would likely lead  to a stronger
mixture of colors in the trajectory map, leading to a predominantly grey hue.

### 2018-08-16 paper/method.tex
beta?

\subsection{Clusters and trajectories}
While the partition of the data into different clusters helps the user to
understand what demographic groups exist and where they are, we are also
interested in the changes in those groups. To this end, we introduce the concept
of \emph{trajectories}, which are composed by regions that are classified into
the same sequence of clusters over the considered period. This enables direct
access to regions that evolved in the same manner. While the interface provides
access to data by census tract, the trajectories are the main unit of
exploration in this work.

@@

\subsection{User interface}
\label{sec:ui}
The initial interface is illustrated in Figure~\ref{fig:ui}. Since demographic
data can be nuanced, with intricate interconnections, we decided against
validating the interface using a synthetic dataset, considering instead data
from the Chicago region between 1970 and 2010, using previous published studies
as corroboration. This region is known for its entrenched racial divide and the
emergence of a \emph{'young urban'} population with a higher education
level~\cite{Delmelle2016,Delmelle2017}. More details about this dataset are
presented in Section~\ref{sec:study}, with an workflow example in
Figure~\ref{fig:chiWorkflow}.

@@

The trajectories overview aims to convey basic information about the
trajectories present in the data, where they are, and what changes are involved.
This is done using three sub panels. The first, on the left, contains a Sankey
diagram illustrating the evolution of the clusters over time. The widths are
proportional to the population involved, the colors follow a policy detailed in
Section~\ref{sec:colors}. A stacked graph could also be used to represent the
proportions of each cluster~\cite{Valdivia2015} with less clutter, since the
transitions between clusters would not be represented. However, this is only
viable if more temporal steps are available, making the plot smoother. Another
option to remove clutter is to remove portions of this plot when trajectories
are selected, but this would change the layout and compromise the user's mental
map.

In our example in Figure~\ref{fig:ui}, we can see that the total population of
Chicago is decreasing. Additionally, the orange and green clusters contain most
of the population and are fairly stable over time. The pink cluster is small and
mostly stable. The purple cluster is increasing, mostly by incorporating areas
that were previously orange. Since the purple group corresponds to the emergent
'young urban' group, this corroborates the findings of
Delmelle~\cite{Delmelle2016,Delmelle2017}. This diagram can also be used to
select specific trajectories, by clicking on the bands, or all trajectories that
contain a specific cluster at a specific time, by clicking on the rectangles.

The next sub panel, in the top middle of Figure~\ref{fig:ui}, is a transition
matrix between the clusters. It indicates the percentage of the population whose
area changed in each pattern. This kind of table can be found in the related
literature~\cite{Delmelle2016}, so it is familiar to the advanced users, and it
not only informs the proportional changes, but allows the selection of the
corresponding trajectories for further analysis.

>>>>

\subsection{Clusters and trajectories}
While the partition of the data into different clusters helps the user to
understand what groups exist and where they are, we are also interested in the
evolution of these groups. We introduce the concept of \emph{trajectories},
composed by regions classified into the same sequence of clusters over the
considered period. This enables direct access to regions that evolved in the
same manner. While the interface provides access to data by individual census
tract, the trajectories are the main unit of exploration in this work.

@@

\subsection{User interface}
\label{sec:ui}
The initial interface is illustrated in Figure~\ref{fig:ui}. Since demographic
data can be nuanced, with intricate interconnections, we decided against
validating the interface using a synthetic dataset, considering instead data
from the Chicago region between 1970 and 2010, using previous published studies
as corroboration. This region is known for its entrenched racial divide and the
emergence of a \emph{'young urban'} population with a higher education
level~\cite{Delmelle2016,Delmelle2017}. More details about this dataset are
presented in Section~\ref{sec:study}.

@@

The trajectories overview aims to convey basic information about the
trajectories, where they are, and what changes are involved. This is done using
three sub panels. The first, on the left, contains a Sankey diagram illustrating
the evolution of the clusters over time. The widths are proportional to the
population involved, the colors follow a policy detailed in
Section~\ref{sec:colors}. A stacked graph could also be used to represent the
proportions of each cluster~\cite{Valdivia2015} with less clutter, since the
transitions between clusters would not be represented. However, this is only
viable if more temporal steps are available, making the plot smoother. Another
option to remove clutter is to remove portions of this plot when trajectories
are selected, but this would change the layout and compromise the user's mental
map.

In our example in Figure~\ref{fig:ui}, the total population of Chicago is
decreasing. Additionally, the orange and green clusters contain most of the
population and are fairly stable over time. The pink cluster is small and mostly
stable. The purple cluster is increasing, mostly by incorporating areas that
were previously orange. Since the purple group corresponds to the emergent
'young urban' group, this corroborates the findings of
Delmelle~\cite{Delmelle2016,Delmelle2017}. This diagram can also be used to
select specific trajectories, by clicking on the bands, or all trajectories that
contain a specific cluster at a specific time, by clicking on the rectangles.

In the next panel, illustrated in the top middle of Figure~\ref{fig:ui}, is a
transition matrix between the clusters. It indicates the percentage of the
population whose area changed between each pair of clusters. This kind of table
can be found in the related literature~\cite{Delmelle2016}, so it is familiar to
the advanced users. It not only informs the proportional changes, but allows the
selection of the corresponding trajectories for further analysis.

### 2018-08-16 paper/study.tex
beta?

While the map of trajectories was mentioned as a "good summary map", how it
related to the clustering method was unclear (C). The methos includes different
options on how the colors are used, but both are sub optimal since reliably
representing several distinct entities using colors is humanly unfeasible.
Indeed, the number of distinguishable colors was a significant constraint, we
found indications that more clusters should be used in some cases, even if eight
clusters is more than what is traditionally considered in these analyses.
Conversely, increasing the number of clusters would also complicate the
interpretation of the results.

The experts also mentioned the poor responsiveness of the method when changes in
the clustering parameters required server-side processing (B,D). Indeed, the
current implementation can take a few minutes to cluster regions with high
number of CTs, like Los Angeles or Brooklyn. Server-processing reduced the
amount of data transfered to client, but it might increase the response time
under load. We implemented a cache policy that greatly improved the performance,
but fully pre-processing the results is not practical due to size of the
parameter space.

>>>>

While the map of trajectories was mentioned as a "good summary map", how it
related to the clustering method was unclear (C). The methods includes different
options on how the colors are used, but both are sub optimal since reliably
representing several distinct entities using colors is humanly unfeasible.
Indeed, the number of distinguishable colors was a significant constraint, we
found indications that more clusters should be used in some cases, even if eight
clusters is more than what is traditionally considered in these analyses.
Conversely, increasing the number of clusters would also complicate the
interpretation of the results.

The experts also mentioned the poor responsiveness of the method when changes in
the clustering parameters required server-side processing (B,D). Indeed, the
current implementation can take a few minutes to cluster regions with high
number of CTs, like Los Angeles or Brooklyn. Server-processing reduced the
amount of data transferred to client, but it might increase the response time
under load. We implemented a cache policy that greatly improved the performance,
but fully pre-processing the results is not practical due to size of the
parameter space.
